%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                 %
%                         ACKNOWLEDGEMENT                         %
%                                                                 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\specialhead{ACKNOWLEDGMENT}

This thesis owes its existence just as much to the support
and collaboration of numerous individuals as it does to
the efforts of the author.
My family must be thanked first for providing so much support
that I was able to give this thesis my full attention.
I must thank Prof. Mark S. Shephard as my mentor
for his unwavering support (and creation) of my career.
My doctoral committee have provided me with many insights:
Prof. Sahni on software development and the needs of advanced
CFD simulations, Prof. Carothers on the ever-changing landscape
of supercomputer architectures, Prof. Anshelevich on graph
theory's elegant solutions to practical problems,
and Prof. Shephard on how topology and adaptivity
can greatly improve existing simulations.

I am honored to have worked alongside students including
Cameron Smith, Brian Granzow, Daniel Zaide, Aleksandr Ovcharenko,
and others whom I consider comrades in the difficult trade
of developing good scientific software.
Other past and present members of SCOREC including Fabien
Delalondre, Seegyoung Seol, Max Bloomfield, and Assad Oberai
have provided guidance in life as well as work.

I admire and thank those who pioneer the adoption of mesh adaptation
in their important simulation workflows including Kenneth Jansen,
Michel Rasquin, Glen Hansen, Chris Kees, and Mike Park.

This thesis is built upon the decades of work by SCOREC including
that of Xiangrong Li, Rao Garimella, and Mark Beall.
It also owes much to the research in mesh adaptivity by
Fr{\'e}d{\'e}ric Alauzet, Adrien Loseille, and Jean-Fran{\c{c}}ois Remacle,
and in message passing by Torsten Hoefler, Andrew Lumsdaine,
and others.
